---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2024_fall")
```

### How to install and load the tidyverse or another library/package

```{r}
install.packages("tidyverse")
library(tidyverse)
```

### How to assign something to a variable
```{r}
number <- 2
number * number
```
###How to load in different types of data files
```{r}
#choose_file_name <- read_insertfiletypehere("insertfilenamehere")
umd_courses <- read_rds("umd_courses.rds")
```

###How to glimpse data
```{r}
glimpse(umd_courses)
#gives a little glimpse at the data so you can get a feel for it
head (umd_courses)
#gives a glimpse at the first 6 rows of data
```
###What is %>% or |>
```{r}
#use ctrl+shift+m to add the figure
#it essentially means "and then do this"
#here's an example
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n()
  )
```

###How to use GROUP_BY, SUMMARISE and ARRANGE
```{r}
#You can do this to group the data and summarize it in a table
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n())

#or you can do this to take all of that and arrange it in a specific way

umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(desc(count_classes))

#you can also group by multiple columns, this makes it so that table shows how many classes each department had each term.
umd_courses |>
  group_by(term, department) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(term, department)

#you can also add more data to the summary like this
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats))

#you can even add summarized data that isn't a count like means or medians
umd_courses |>
  group_by(term) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    mean_seats = mean(seats),
    median_seats = median(seats))
```

