---
title: "week7_recap"
author: "Daniel Trielli"
date: "2023-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load Data

Run the codeblock below to load the data.

```{r}
earthquakes <- read_csv('https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv')

#Setting time column as datetime
earthquakes <- earthquakes |> mutate(time = as_datetime(time))
```

#### Answer the questions below

Most questions have a code block and a space for an answer below. Write the code you think is necessary and, in the answer space, write out what you did and what was the result.

------------------------------------------------------------------------

#### **Q1** Do you notice anything different with this read_csv()? Particularly in the file name?

**A1:The read_csv is linked to a url rather than a file.**

------------------------------------------------------------------------

#### **Q2** How many records there are there in this dataset? What do they mean and what useful information we can gather from it, looking at the columns?

**A2: There are 9139 records in this dataset. Each one represents every earthquake/seismic incident that has occurred globally in 2025. We can gather when and where earthquakes occur, which means we can determine how frequently earthquakes occur in different places and at what scale. We can also see which types of tremors occur.**

------------------------------------------------------------------------

#### **Q3** How do I reorganize this list to see the ones that are the deepest first? What is the depth that shows up for the deepest one, and its magnitude?

```{r}
earthquakes %>% 
  arrange(desc(depth))
```

**A3: So far this year, the deepest earthquake occured on Oct. 4 and was 643km deep with a magnitude of 4.2.**

------------------------------------------------------------------------

#### **Q4** I just want to see the earthquakes with a magnitude larger than 6. How do I do that? And how many are there that fit this criteria?

```{r}
six_eqs <- earthquakes %>% 
  filter(mag > 6)
```

**A4: There are 12 earthquakes with a maginitude over 6.0 that occurred in 2025.**

------------------------------------------------------------------------

#### **Q5** What about if I want to see earthquakes that have both a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [both]{.underline} these criteria?

```{r}
sixandtwenty_eqs <- earthquakes %>% 
  filter(mag > 6 & depth < 20)
```

**A5: There are eight earthquakes that have both a magnitude larger than 6 and a depth smaller than 20km.**

------------------------------------------------------------------------

#### **Q6** What about if I want to see earthquakes that either have a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [either]{.underline} these criteria?

```{r}
sixortwenty_eqs <- earthquakes %>% 
  filter(mag > 6 | depth < 20)
```

**A6: The number increases dramatically with 7608 earthquakes that either have a magnitude larger than 6 and a depth smaller than 20km.**

------------------------------------------------------------------------

#### **Q7** I'm interested in finding earthquakes that took place in Alaska. Which column should I look at? How do I use it to find all the earthquakes in Alaska? How many earthquakes took place there?

```{r}
ak_earthquakes <- earthquakes %>% 
  filter(str_detect(place, "Alaska"))
```

**A7: You should look at the "place" column. 2355 earthquakes took place there this year.**

------------------------------------------------------------------------

#### **Q8** I notice that there is a column called 'type', that seems to have different kinds of tremors. What is the best way to find what are all the possible types of tremors, and counting how many of each there are in this data set? What are the first two most common types of tremors in this data set?

```{r}
tremor_types <- earthquakes %>%
  group_by(type) %>% 
  summarize(count=n()) %>% 
  arrange(desc(count))
```

**A8: The first two most common types of tremors are earthquakes, 8936, and quarry blasts, 127.**

------------------------------------------------------------------------

#### **Q9** How do I calculate the average depth of each type of tremor? What is the average depth of the earthquake type in this data set? Is there anything that seems unusual or surprising in this finding?

```{r}
tremor_depths <- earthquakes %>%
  group_by(type) %>%
  summarize(avg_depth = mean(depth))
```

**A9: I thought it was interesting that the landslide had no depth to it. I also found it interesting that some of the average depths were negative since you normally think of tremors as being underground.**

------------------------------------------------------------------------

#### **Q10** I'm interested, in the future, to see the hours in which earthquakes happen. How can I extract the hour from the time column?

```{r}
eq_hours <- earthquakes %>% 
  mutate(time = ymd_hms(time),
         hour_only= hour(time))
```

**A10: use lubridate as seen above**

------------------------------------------------------------------------

#### **Q11** I want to make a note of all the records in this data set that I consider serious. For my purposes, I'm thinking that all tremors that have a magnitude that is larger than 3 are serious. How do I automatically create a new column making that distinction?

```{r}
eq_3plus <- eq_hours %>% 
  mutate(
    serious = case_when(
        mag > 3 ~ "yes",
        .default = "no"
      )
  )
```

**A11: View code above**

------------------------------------------------------------------------

#### **Q12** I have no idea how earthquakes work and I'm interested in seeing if there is a particular time of day in which serious earthquakes happen. How can I see that condensed in a table with all the hours in a day and all the serious earthquakes in each hour? What is the hour with fewer serious earthquakes and the one with the most serious earthquakes?

```{r}
hourly_serious_eqs <- eq_3plus %>% 
  filter(serious == "yes") %>% 
  group_by(hour_only) %>% 
  summarize(total_hourly_serious_eqs = n()) %>% 
  arrange(desc(total_hourly_serious_eqs))
```

**A12**: The hours with the most earthquakes are 1-2 a.m., and 4 p.m. to 6 p.m. The hour with the least is 12 a.m. to 1 a.m.

------------------------------------------------------------------------

#### EXTRA Q What other questions are you interested in and how would you ask them in R?


```{r}
#How often do they have to review the earthquake entries? Entries have to be reviewed more than twice as much as they don't need to be.
eq_reviews <- eq_hours %>% 
  group_by(status) %>% 
  summarize(total = n())
```

```{r}
#How many of these are international or in the US? Almost all of the entries occur in the US.
internatl_vs_us <- eq_hours %>% 
  mutate(
    global_status = case_when(
        str_detect(place, "CA") ~ "US",
        str_detect(place, "Alaska") ~ "US",
        str_detect(place, "Washington") ~ "US",
        str_detect(place, "New Mexico") ~ "US",
        str_detect(place, "Wyoming") ~ "US",
        str_detect(place, "Texas") ~ "US",
        str_detect(place, "California") ~ "US",
        str_detect(place, "Missouri") ~ "US",
        str_detect(place, "Oklahoma") ~ "US",
        str_detect(place, "Arizona") ~ "US",
        str_detect(place, "Alabama") ~ "US",
        str_detect(place, "Arkansas") ~ "US",
        str_detect(place, "Colorado") ~ "US",
        str_detect(place, "Connecticut") ~ "US",
        str_detect(place, "CT") ~ "US",
        str_detect(place, "Delaware") ~ "US",
        str_detect(place, "Florida") ~ "US",
        str_detect(place, "Georgia") ~ "US",
        str_detect(place, "Hawaii") ~ "US",
        str_detect(place, "Idaho") ~ "US",
        str_detect(place, "Illinois") ~ "US",
        str_detect(place, "Indiana") ~ "US",
        str_detect(place, "Iowa") ~ "US",
        str_detect(place, "Kansas") ~ "US",
        str_detect(place, "Kentucky") ~ "US",
        str_detect(place, "Louisiana") ~ "US",
        str_detect(place, "Maine") ~ "US",
        str_detect(place, "Maryland") ~ "US",
        str_detect(place, "MA") ~ "US",
        str_detect(place, "Massachusetts") ~ "US",
        str_detect(place, "Michigan") ~ "US",
        str_detect(place, "Minnesota") ~ "US",
        str_detect(place, "Mississippi") ~ "US",
        str_detect(place, "MS") ~ "US",
        str_detect(place, "Missouri") ~ "US",
        str_detect(place, "Montana") ~ "US",
        str_detect(place, "Nebraska") ~ "US",
        str_detect(place, "Nevada") ~ "US",
        str_detect(place, "New Hampshire") ~ "US",
        str_detect(place, "NH") ~ "US",
        str_detect(place, "New Jersey") ~ "US",
        str_detect(place, "New York") ~ "US",
        str_detect(place, "NC") ~ "US",
        str_detect(place, "North Carolina") ~ "US",
        str_detect(place, "North Dakota") ~ "US",
        str_detect(place, "ND") ~ "US",
        str_detect(place, "Ohio") ~ "US",
        str_detect(place, "Oregon") ~ "US",
        str_detect(place, "Pennsylvania") ~ "US",
        str_detect(place, "PA") ~ "US",
        str_detect(place, "Rhode Island") ~ "US",
        str_detect(place, "RI") ~ "US",
        str_detect(place, "South Carolina") ~ "US",
        str_detect(place, "SC") ~ "US",
        str_detect(place, "South Dakota") ~ "US",
        str_detect(place, "SD") ~ "US",
        str_detect(place, "Tennessee") ~ "US",
        str_detect(place, "TN") ~ "US",
        str_detect(place, "Utah") ~ "US",
        str_detect(place, "Vermont") ~ "US",
        str_detect(place, "Virginia") ~ "US",
        str_detect(place, "West Virginia") ~ "US",
        str_detect(place, "WV") ~ "US",
        str_detect(place, "Wisconsin") ~ "US",
        str_detect(place, "Guam") ~ "US",
        str_detect(place, "Puerto Rico") ~ "US",
        str_detect(place, "PR") ~ "US",
        str_detect(place, "American Samoa") ~ "US",
        str_detect(place, "Northern Mariana Islands") ~ "US",
        str_detect(place, "U.S.") ~ "US",
        str_detect(place, "D.C.") ~ "US",
        str_detect(place, "DC") ~ "US",
        str_detect(place, "District of Columbia") ~ "US",
        .default = "international"
      ))
    
global_status_df <- internatl_vs_us %>% 
    group_by(global_status) %>% 
    summarize(total = n())
```

```{r}
#How long does it take to update the data? It takes more than four times as long for them to update entires for earthquakes with a magnitude greater than 3.
updated_time_df <- eq_3plus %>% 
  mutate(update_time = updated - time)

updated_time_df <- updated_time_df %>%
  group_by(serious) %>% 
  summarize(avg_update_time = mean(update_time))
```
