---
title: "lab_03"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)
# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=369008)

# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")

# add underscores to names and lowercase everything
cleaned_md_grants_loans <- md_grants_loans |>
  clean_names() %>% 
  mutate(grantor = str_to_lower(grantor), grantee = str_to_lower(grantee), description = str_to_lower(description))

cleaned_payments <- payments |>
  clean_names() %>% 
  mutate(agency_name = str_to_lower(agency_name), vendor_name = str_to_lower(vendor_name), category = str_to_lower(category))

```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1: There are 70 records going to Prince George's County Public Schools.

```{r}
pg_schools <- cleaned_payments |>
  filter(vendor_name == "prince georges county public schools" | vendor_name == "prince george's county public schools" | vendor_name == "pg county public school" | vendor_name == "prince georges county public school" | vendor_name == "prince georges public schools" | vendor_name == "prince georges county school")

nrow(pg_schools)
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: 2021 had the largest total amount accepting \$1,470,133,533 in payments. I don't know much about school budgets but that seems like a huge number for one county. However, my inexperience with school budgets is backed up by all of the other fiscal year amounts being somewhat similar. The number is still a whole 100,000,000 greater than the next largest sum of payments to the school district, which is odd. I'm also thrown off by the fact that there are eight whole years missing from the sums: 2011, 2012, 2016, 2022, 2023, and 2024. It makes me wonder if there are just things mislabeled or if there are truly gaps that large in the dataset.

```{r}
pg_schools |>
  group_by(fiscal_year) |>
  summarise(amount_schools = sum(amount)) |>
  arrange(desc(amount_schools))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. Here are some examples of vendors that are similar and should be combined: whiting turner contracting co and whiting turner contracting company\
plano-condon, llc and plano-coudon llc\
ballinger and the ballinger co

You would use the case_when() function to sift through all of the vendor_name values and search for terms within the values, like "ballinger," and reassign them a common name.

```{r}
umd_capital_projects <- cleaned_payments |>
  filter(agency_name == "capital projects- university of maryland")

umd_capital_projects |>
  group_by(vendor_name) |>
  summarise(umd_amount = sum(amount)) |>
  arrange(desc(umd_amount))
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: Stem cell research funding has greatly changed over time with the earliest year available in the data (2009) and the latest (2023) being the years that saw the greatest investment in stem cell research. In between, funding has largely floated around $7 million with frequent dips and rises.

```{r}
stem_cell_research <- cleaned_md_grants_loans |>
  filter(description == "stem cell research grant" | description == "stem cell research grant, pre-commercialization research")

stem_cell_research |>
  group_by(fiscal_year) |>
  summarise(stem_cell_amount = sum(amount)) |>
  arrange(desc(fiscal_year))
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: You would have to standardize all of the entries that reference the University of Maryland (ex: "university of md college park" or "university of maryland") using the case_when() function to sift through all of the grantee values and search for terms that reference UMD and reassign them a common name.

"pregnancy aid centers, inc." was an immediate standout because it references a women's health center in College Park that is receiving grants/loans from the state. "elizabeth gallauresi" is also just a singular person which draws the question of who that is and why they're getting grants from the state. After looking it up I was still confused because she works for the Silver Spring branch of the Maryland State Arts Council, so I'm almost wondering if she's having state grants sent to her home address or an alternative office.


```{r}
umd_located_grants <- cleaned_md_grants_loans |>
  filter(zip_code == 20742 | zip_code == 20740)

umd_located_grants |>
  group_by(grantee) |>
  summarise(umd_grantee_amt = sum(amount)) |>
  arrange(desc(umd_grantee_amt))
```
